---
output: 
  pdf_document:
    fig_width: 5
    fig_height: 4
    fig_caption: false
---
## Wealth of Nations: Analysis of Income Disparity and Relationship to Poverty
#### Author: Nirmal Labh  [*HUID: 80625529*]
#### Contact TF: Kela Roberts 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(pander)
library(ggplot2)
require(WDI)
```


```{r echo=FALSE}
options(warn=-1)
options(scipen=999)
panderOptions('table.split.table', Inf)
panderOptions('table.split.cells', Inf) 
panderOptions('table.alignment.default', 'left') 
```

```{r echo=FALSE}
raw_country_code = read.csv('C:/STAT-100/Project/country_codes.csv',skip=0,header = FALSE)
country_code=c(raw_country_code[1])
all_countries=as.character(country_code$V1)

# Zambia country code of 'NA' is read in as R NA
# Updating it to be character 'NA'
all_countries[which(is.na(all_countries))]='NA'

# Get WDI data using API and functions as part of CRAN WDI module
wdi_data=WDI(country = all_countries, indicator=c('BN.GSR.FCTY.CD','NY.GDP.MKTP.CD','SI.POV.NAHC','SI.DST.FRST.10','SI.DST.10TH.10','SE.ADT.LITR.ZS','SE.ADT.1524.LT.ZS','SP.POP.TOTL'),start=1970,end=2013,extra=FALSE)
names(wdi_data)=c('isoc2','country','year','net_income','gdp','poverty','income_low10','income_high10','literacy_a','literacy_y','population')

# 2 literacy pecentages - under 15 and above 15 - being averaged
# as proportion of population in each category i snot known
wdi_data$literacy=(wdi_data$literacy_a+wdi_data$literacy_y)/2

```

```{r echo=FALSE}
# average income share of top 10% for year 2003 and 2013 
mean_hi10_2003=round(mean(wdi_data[which(!is.na(wdi_data$income_high10) & wdi_data$year=='2003'),'income_high10']),0)
mean_hi10_2013=round(mean(wdi_data[which(!is.na(wdi_data$income_high10) & wdi_data$year=='2013'),'income_high10']),0)

# average income share of top 10% for year 2003 and 2013
mean_low10_2003=round(mean(wdi_data[which(!is.na(wdi_data$income_low10) & wdi_data$year=='2003'),'income_low10']),2)
mean_low10_2013=round(mean(wdi_data[which(!is.na(wdi_data$income_low10) & wdi_data$year=='2013'),'income_low10']),2)

# average GDP for year 2003 and 2013
mean_gdp_2003=round(mean(wdi_data[which(!is.na(wdi_data$gdp) & wdi_data$year=='2003'),'gdp'])/1000000000,0)
mean_gdp_2013=round(mean(wdi_data[which(!is.na(wdi_data$gdp) & wdi_data$year=='2013'),'gdp'])/1000000000,0)

# average poverty for year 2013
mean_poverty_2013=round(mean(wdi_data[which(!is.na(wdi_data$poverty) & wdi_data$year=='2013'),'poverty']),0)
```


## Abstract


It is widely understood that income disparity has continued to increase. While world average GDP more than doubled from `r mean_gdp_2003` billion in 2003 to `r mean_gdp_2013` billion in 2013, and the average share of income for the higher 10% went up from `r mean_hi10_2003`% to `r mean_hi10_2013`%, the income of the lower 10% hardly increased from `r mean_low10_2003`% to `r mean_low10_2013`%. Average world poverty stood at `r mean_poverty_2013`%.

To attain better social stability, it is vital to improve the income of the less fortunate ones reflected in the 10% lower income population as well as reduce the headcount of people living under the poverty lines.

Using simple and multiple linear regressions on income data for all countries between 1974 and 2013 obtained from World Bank^1^, we investigate below three explanatory variables as predictors of poverty level:

* GDP
* Income level of lower and higher 10%
* Literacy

The *model* as well as the *three explanatory variables* have extremely *small p-value*.

The paper therefore concludes that GDP growth, coupled with increase in literacy has important implications for improving income of low earners as well as reducing the level of poverty.



## I. Introduction


Every country strives to increase GDP, a reflection of increased economic activity, and hence of increased productivity, trade and employment among many other things. It has direct consequences in improving income levels and poverty reduction that we look at in this paper.

We start off by looking at below variables in the WDI data^1,2^:

* Quantitative variables:
    + Income of lower 10%
    + Income of higher 10%
    + Wealth i.e. Net Income
    + Poverty Level
    + Population Count
    
* All of the data has below two dimensions:
    + Country: *While widely accepted count of countries is 195, it lists 214, more so as distinct data sources*
    + Year: *The data is for years starting 1974 to 2013*


In this paper we attempt to evaluate and validate correlation between:

* income share of the lower 10% and the higher 10%
* poverty and income share of higher 10%
* poverty and income share of lower 10%.
* poverty and literacy 
 

Vizualization of the wealth and income disparity as well as the above four listed correlations is being done in R using *ggplot2*^3^ CRAN module.


## II. Methods

The data used in the analysis are retrieved using the *WDI CRAN module*^2^. This module provides direct API access to the World Bank^1^ data from within R code. All of the numbers and charts appearing in this paper have been prepared using R markdown. Complete .Rmd file for this paper is available at [github](https://github.com/nlabh01/STAT-100-Project). R code for calculations and charts are included in the *Appendix* section.


Data Description:

* WDI Indicators used to lookup specific data points for the study:

```{r echo=FALSE}
indicator_description=WDIsearch(string='BN.GSR.FCTY.CD$|NY.GDP.MKTP.CD$|SI.POV.NAHC$|SI.DST.FRST.10$|SI.DST.10TH.10$|SE.ADT.LITR.ZS$|SE.ADT.1524.LT.ZS$|SP.POP.TOTL$', field='indicator')
colnames(indicator_description)=c('Indicator','Description')

set.caption('Subset of WDI Indicators')
pander(indicator_description)
```


* Sample Data:
```{r echo=FALSE}
wdi_data_disp=wdi_data
wdi_data_disp$gdp=wdi_data_disp$gdp/1000000000
names(wdi_data_disp)=c('isoc2','Country','Yr','net_income','GDP','Poverty','IncmLo10','IncmHi10','literacy_a','literacy_y','population','Literacy')

set.caption('Sample data based on Indicators from Table 1')
pander(head(wdi_data_disp[which( !is.na(wdi_data_disp$Literacy) & !is.na(wdi_data_disp$IncmLo10) ),c(2,3,5,6,7,8,12)]))

```

We start off by examining the simple correlation between the below pairs:

  * Income of lower 10% share vs income of higher 10% share (section III.1.)
  * Poverty vs Higher 10% Income Share (section III.2)
  * Poverty vs Lower 10% Of Income Share (section III.3)
  * Povery and Literacy (section III.4)

Establishing there is correlation, we further look at their combinations in the form of Multiple Regression Model.
With iteratibve refinement, we com up to the model in section III.5, verify the distributions of residuals to be normal in section III.6, and continue on to do additional verification using ANOVA in section III.7. We see very good *p-values* that are nearly zero for this model.

For completeness, one of the models that was rejected is listed in Appendix - section VI.12. It was rejected on the basis that most of the *p-values* associatd with the explanatory variables were rather big and hence not significant.


## III. Results


###1. Income Disparity

Based on the data, scatter plot with *income of lower 10% share on the y-axis* and *income of higher 10% share* on the x-axis shows *strong negative linear relationship* between the two variables. 

It validates commonly held notion that *rich are getting richer and poor are getting poorer*.

```{r echo=FALSE}
ggplot(wdi_data, aes(x=income_high10, y=income_low10)) + geom_point(size=1) +  geom_smooth(method="lm") + 
  labs(x='Income Held By Higher 10%', y='Income Held By Lower 10%')
```

We further validate this strong negative correlation, notice the nearly zero *p-value* and *negative co-efficient*:

```{r echo=FALSE}
summary(lm(income_low10 ~ income_high10, data=wdi_data))
```

Word of caution here is that the relationship seems to be very slightly curved, though mostly linear. We may need to consider additional variables as confounders that can then help explain this relationship better. Eliminating extreme outliers is yet another way we can deal with linearity.


###2. Poverty and Higher 10% Of Income Share

Below we see *positive correlation between poverty and income share of higher 10%*.

This is indicating that *poverty level only rises when the higher 10% income holders have even more income*.


```{r echo=FALSE}
ggplot(wdi_data, aes(x=income_high10, y=poverty)) + geom_point(size=1) +  geom_smooth(method="lm") + 
  labs(x='Income Held By Higher 10%', y='Poverty')
```

Looking at the simple regression model, we find fairly positive correlation, notice the nearly zero *p-value* with positive co-efficient below. 

```{r echo=FALSE}
summary(lm(poverty ~ income_high10, data=wdi_data))
```


###3. Poverty and Lower 10% Of Income Share

Below we see *negative correlation between poverty and income share of lower 10%*.

This is indicating that *poverty level goes down as the income goes up for the ones holding lower 10% of the income*.


```{r echo=FALSE}
ggplot(wdi_data, aes(x=income_low10, y=poverty)) + geom_point(size=1) +  geom_smooth(method="lm") + 
  labs(x='Income Held By Lower 10%', y='Poverty')
```

Looking at the simple regression model, we find fairly negative correlation, notice the nearly zero *p-value* with negative co-efficient below. 


```{r echo=FALSE}
summary(lm(poverty ~ income_low10, data=wdi_data))
```


###4. Povery and Literacy

Below we see *negative correlation between poverty and literacy*. We would expect literacy can get people to get more and/or better employment. So, **bringing literacy to uneducated parts of the population can help reduce poverty**.


```{r echo=FALSE}
ggplot(wdi_data, aes(x=literacy, y=poverty)) + geom_point(size=1) +  geom_smooth(method="lm") + 
  labs(x='Literacy', y='Poverty')
```

The trend is somewhat questionable in the statistical sense, in that the data seems to be fanning out.
Further analysis using the regression model indicates low *p-value*, but not a very high *R^2^*.
So, we take into account the combination of all of the above variables to proceed with analysis using multiple regression.

```{r echo=FALSE}
summary(lm(poverty ~ income_low10, data=wdi_data))
```



###5. Multiple Regression Model

While we observe relationships above in four cases using simple regression model, we continue on to analyze them in a combined way using multiple regression.

We attempt to predict poverty using explanatoray varaibles as *gdp*, *income_low10* and *literacy*.
Iterations of other combinations were dropped using stepwise refinment of the model.

```{r}
lmWDI=lm(poverty ~ gdp + income_low10 + literacy, data = wdi_data)
summary(lmWDI)
```


###6. Distribution of Residuals

We check for residuals to see if they are fairly *normally distributided* below. We find that the residuals for the model more or less does follow normal distribution. However, we do see some extreme outliers, masking the symmetry of the distribution to some extent.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# histogram of residuals
flm=fortify(lmWDI)
ggplot(flm, aes(x=.resid)) + geom_histogram(fill="grey", color="black")
```



###7. Analysis of Variance

Here, we observe very small p-values.

```{r}
anova(lmWDI)
```


## IV. Conclusion and Discussion

This paper demonstrates the poverty level and income share of lower 10% is significantly correlated to GDP, income of lower 10%, and literacy, with *p-values* for each of the three explanatory variables close to zero. With such *small p-values*, *R^2* of *41%* , *F-statistic* of *28.7* and a fairly *normal distribution of the residuals*, we have significant confidence in the model.

To further read into this model and reason out, we can say that increase in GDP resulting from many factors such as increased productivity, trade, etc causes increase in employment. More economic activities and increase in employment directly as well as indirectly increases income of wage earners. That in turn results in an uplift to the earnings of the lower income population and help reduce poverty.

One flaw in this study is that there are too many possibile confounding variables that can skew the model. Also, all of *the study is inherently observational in nature* and so *no conclusions of cause-and-effect should be drawn*. As frequently stated, *correlation is not causation*. So then, what good is this study for? Well, it is good in that it adds value by studying and observing different economic and social attributes that affect income and hence betterment for the population, in particular the ones that are in the lower income or living in poverty.

The wider implication of this study is that government bodies and organizations should recognize the critical importance of the *growing income disparity*, its *negative effect on poverty*, and *potentially positive effect of increase in literacy*.

This study is in no way conclusive and *can be improved with injection of additional confounding variables such as gender, region, race, geo-political factors* and many more. It may be worthwhile to investigate such additional factors and statistical methods to analyze and potentially improve income for low earners and poverty reduction.


## V. References

1. World Bank: Data Catalog - Wealth Accounting
   * <http://data.worldbank.org/data-catalog/wealth-accounting>
   
2. WDI: World Development Indicators (World Bank)
   * *Search, extract and format data from the World Bank's World Development Indicators in R*
   * <http://cran.r-project.org/web/packages/WDI/WDI.pdf>
   
3. ggplot2: An implementation of the Grammar of Graphics
  * <http://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf>
  
4. pander: An R Pandoc Writer
   * <http://cran.r-project.org/web/packages/pander/pander.pdf>
   

## VI. Appendix


###1. Various CRAN Modules and R Options Used

```{r echo=TRUE, message=FALSE, warning=FALSE}
require(WDI)
library(ggplot2)
library(pander)


options(warn=-1)
options(scipen=999)
panderOptions('table.split.table', Inf)
panderOptions('table.split.cells', Inf) 
panderOptions('table.alignment.default', 'left') 
```

###2. Acquiring WDI Data

```{r}
# initial extract of all countries using WDI(country = "all", indicators...) had all 
# countries as well as additional rows for regional aggregations. Extracted the 
# country codes, removed the region rows, then saved it to local CSV file
raw_country_code = read.csv('C:/STAT-100/Project/country_codes.csv',
                            skip=0,header = FALSE)
country_code=c(raw_country_code[1])
all_countries=as.character(country_code$V1)

# Zambia country code of 'NA' is read in as R NA
# Updating it to be character 'NA'
all_countries[which(is.na(all_countries))]='NA'

# Get WDI data using API and functions as part of CRAN WDI module
wdi_data=WDI(country = all_countries, indicator=c('BN.GSR.FCTY.CD','NY.GDP.MKTP.CD',
                                                  'SI.POV.NAHC','SI.DST.FRST.10',
                                                  'SI.DST.10TH.10','SE.ADT.LITR.ZS',
                                                  'SE.ADT.1524.LT.ZS','SP.POP.TOTL'),
             start=1970,end=2013,extra=FALSE)
names(wdi_data)=c('isoc2','country','year','net_income','gdp','poverty',
                  'income_low10','income_high10','literacy_a','literacy_y',
                  'population')

# 2 literacy pecentages - under 15 and above 15 - being averaged
# as proportion of population in each category i snot known
wdi_data$literacy=(wdi_data$literacy_a+wdi_data$literacy_y)/2

```


###3. Various mean(s) used in the paper

```{r}
# average income share of top 10% for year 2003 and 2013 
mean_hi10_2003=
  round(mean(wdi_data[which(!is.na(wdi_data$income_high10) & wdi_data$year=='2003'),
                      'income_high10']),0)
mean_hi10_2013=
  round(mean(wdi_data[which(!is.na(wdi_data$income_high10) & wdi_data$year=='2013'),
                      'income_high10']),0)

# average income share of top 10% for year 2003 and 2013
mean_low10_2003=
  round(mean(wdi_data[which(!is.na(wdi_data$income_low10) & wdi_data$year=='2003'),
                      'income_low10']),2)
mean_low10_2013=
  round(mean(wdi_data[which(!is.na(wdi_data$income_low10) & wdi_data$year=='2013'),
                      'income_low10']),2)

# average GDP for year 2003 and 2013
mean_gdp_2003=
  round(mean(wdi_data[which(!is.na(wdi_data$gdp) & wdi_data$year=='2003'),'gdp'])
        /1000000000,0)
mean_gdp_2013=
  round(mean(wdi_data[which(!is.na(wdi_data$gdp) & wdi_data$year=='2013'),'gdp'])
        /1000000000,0)

# average poverty for year 2013
mean_poverty_2013=
  round(mean(wdi_data[which(!is.na(wdi_data$poverty) & wdi_data$year=='2013'),
                      'poverty']),0)
```


###4. Indicators Used

```{r}
# WDI Indicators used to lookup specific data points for the study:
indicator_description=WDIsearch(string='BN.GSR.FCTY.CD$|NY.GDP.MKTP.CD$|SI.POV.NAHC$|
                                SI.DST.FRST.10$|SI.DST.10TH.10$|SE.ADT.LITR.ZS$|
                                SE.ADT.1524.LT.ZS$|SP.POP.TOTL$', field='indicator')
colnames(indicator_description)=c('Indicator','Description')

set.caption('Subset of WDI Indicators')
pander(indicator_description)
```


###5. Sample Data

```{r}
# Sample Data:
wdi_data_disp=wdi_data
wdi_data_disp$gdp=wdi_data_disp$gdp/1000000000
names(wdi_data_disp)=c('isoc2','Country','Yr','net_income','GDP','Poverty','IncmLo10',
                       'IncmHi10','literacy_a','literacy_y','population','Literacy')

set.caption('Sample data based on Indicators from Table 1')
pander(head(wdi_data_disp[which( !is.na(wdi_data_disp$Literacy) 
                                 & !is.na(wdi_data_disp$IncmLo10) ),
                          c(2,3,5,6,7,8,12)]))

```


###6. Income Disparity with the Simple Regression Model

```{r fig.width=4, fig.height=3}
# Income Disparity:
ggplot(wdi_data, aes(x=income_high10, y=income_low10)) + 
  geom_point(size=1) +  geom_smooth(method="lm") + 
  labs(x='Income Held By Higher 10%', y='Income Held By Lower 10%')

# model for income disparity
summary(lm(income_low10 ~ income_high10, data=wdi_data))
```


###7. Povery and Higher 10% Of Income Share with the Simple Regression Model

```{r fig.width=4, fig.height=3}
# Poverty and Higher 10% Of Income Share
ggplot(wdi_data, aes(x=income_high10, y=poverty)) + geom_point(size=1) +  
  geom_smooth(method="lm") + labs(x='Income Held By Higher 10%', y='Poverty')

summary(lm(poverty ~ income_high10, data=wdi_data))
```


###8. Povery and Lower 10% Of Income Share with the Simple Regression Model

```{r fig.width=4, fig.height=3}
# Poverty and Lower 10% Of Income Share
ggplot(wdi_data, aes(x=income_low10, y=poverty)) + geom_point(size=1) +  
  geom_smooth(method="lm") + labs(x='Income Held By Lower 10%', y='Poverty')

# simple regression model
summary(lm(poverty ~ income_low10, data=wdi_data))
```


###9. Povery and Literacy with the Simple Regression Model

```{r fig.width=4, fig.height=3}
# Povery and Literacy
ggplot(wdi_data, aes(x=literacy, y=poverty)) + 
  geom_point(size=1) +  geom_smooth(method="lm") + labs(x='Literacy', y='Poverty')

# simple regression model
summary(lm(poverty ~ income_low10, data=wdi_data))
```


###10. Multiple Regression Model and Residual Histogram

```{r message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
# Multiple Regression Model
lmWDI=lm(poverty ~ gdp + income_low10 + literacy, data = wdi_data)
summary(lmWDI)

# histogram of residuals
flm=fortify(lmWDI)
ggplot(flm, aes(x=.resid)) + geom_histogram(fill="grey", color="black")
```

```


###11. Analysis of Variance

```{r}
# Analysis of Variance
anova(lmWDI)
```


###12. One of the alternative models among the few that were looked at

```{r}
# Alternative model among the many looked at
summary(lm(poverty ~ gdp + income_low10 + literacy + population + 
             net_income + income_high10, data = wdi_data))
```
